\chapter{Typological overview}
\label{chap:2}

This section is a~brief overview covering the most central features of Palula. For more in"=depth coverage of each topic, and for information on those not explicitly covered here, the later chapters will need to be consulted.


\section{Phonology}
\label{sec:2-1}


With its 27--30 members, the Palula consonant inventory is moderately large \citep[10--13]{maddieson2005a} and would be considerably larger if aspiration were segmentally analysed, in which case another 20 members would be added to the inventory. There are five basic places of articulation (labial, dental, retroflex, palatal and velar), with a~voicing contrast in the plosive and fricative sets: 


\begin{table}[H]
\begin{tabular}{ l@{\hspace{20pt}} l@{\hspace{20pt}} l@{\hspace{20pt}} l@{\hspace{20pt}} l@{\hspace{20pt}} l@{\hspace{20pt}} l@{\hspace{20pt}} }
/p &
t &
ʈ &
&
k &
(q) &
\\
b &
d &
ɖ &
&
ɡ &
&
\\
&
ʦ &
ʈʂ &
ʨ &
&
&
\\
(f) &
s &
ʂ &
ɕ &
x &
&
h\\
&
z &
(ʐ) &
ʑ &
ɣ &
&
\\
m &
n &
ɳ &
&
&
&
\\
&
r &
ɽ &
&
&
&
\\
&
l &
&
&
&
&
\\
w &
&
&
j/ &
&
&
\\
\end{tabular}
\end{table}


Note that aspiration occurs extensively in Palula, but is analysed as a~property of the word (see below).


Palula has ten phonemic vowels, comprising five basic qualities, each having a~long and a~short counterpart. This inventory forms a~symmetrical and typologically common system: 

\begin{center}
\begin{tabular}{ll@{\hspace{20pt}}ll@{\hspace{20pt}}ll@{\hspace{20pt}}ll@{\hspace{20pt}}ll}
/iː &
i &
&
&
&
&
&
&
uː &
u\\
&
&
eː &
e &
&
&
oː &
o &
&
\\
&
&
&
&
aː &
a/ &
&
&
&
\\
\end{tabular}
\end{center}



Vowel nasalisation is a~marginal, possibly emerging, feature in the language, but not so far fully contrastive.


The language has a~moderately complex syllable structure \citep[54--57]{maddieson2005b}, permitting a~limited number of consonant combinations before or after the vowel nucleus. Although the final clusters are slightly more permissive, there is a~tendency to drop the final consonant in word final clusters:


\begin{table}[H]
\begin{tabularx}{\textwidth}{ l l l l}
Before V: &
C + /r/ &
/kraːm/ &
`work'\\
&
C + approximant &
/swaːnu/ &
`is sleeping'\\
After V: &
nasal + C &
/ɡʰreːɳɖ/ &
`knot'\\
&
fricative + plosive (/s + t/, /ʂ + ʈ/) &
/ɡʰoːʂʈ/ &
`house'\\
&
C+ /r/ (/t + r/) &
/putr/ &
`son'\\
\end{tabularx}
\end{table}


Aspiration is analysed as a~property of the word as a~whole in Palula. The feature, realised as [ʰ] with a~voiceless consonant and as [[ʱ]] with a~voiced consonant (usually concomitant with the following vowel having breathy voice), occurs only once in a~word, and is represented with h in the general transcription in this work. Some minimal pairs illustrate the contrastiveness of this feature: 


\begin{table}[H]
\begin{tabularx}{\textwidth}{ l l@{\hspace{20pt}} l l }
/bʰoːla/ &
`were able to' &
/boːla/ &
`hair'\\
/kʰareːɽi/ &
`bolt' &
/kareːɽi/ &
`leopardess'\\
/wʰiː/ &
`he/she will come down' &
/wiː/ &
`water'\\
\end{tabularx}
\end{table}


Main stress falls on the final or the penultimate syllable of the lexical root. The stressed
syllable receives pitch accent, phonetically realised as: a) high level or falling on a~short vowel
[\'{}], represented in this work as \textit{á} (in polysyllabic words, elsewhere no marking); b)
rising on a~long vowel [\v{}], represented \textit{aá}; or c) falling on a~long vowel [\^{}],
represented as \textit{áa}. Pitch accent is contrastive:


\begin{table}[H]
\begin{tabularx}{\textwidth}{ l l Q l l Q l l Q }
/seːtí/ &
\textit{seetí} &
`looked after' &
/sêːti/ &
\textit{séeti} &
`thigh' &
&
&
\\
/děːdi/ &
\textit{deédi} &
`burnt \textsc{f}' &
/dêːdi/ &
\textit{déedi} &
`grand\-mother' & & &
\\
/hár/ &
\textit{har} &
`every' &
/hǎːr/ &
\textit{haár} &
`defeat' &
/hâːr/ &
\textit{háar} &
`take away!' \\
\end{tabularx}
\end{table}


\section{Morphology}
\label{sec:2-2}

Palula morphology is suffixing, and formatives are almost exclusively concatenative \citep[86--9]{bickelnichols2005a}, with a~moderately high degreee of synthesis \citep[94--97]{bickelnichols2005c}. 


Nouns are inflected for number (singular, plural) and case (nominative, oblique, genitive). In most of the declensional classes nominative plural and oblique singular are cumulated into a~single formative \citep[90--93]{bickelnichols2005b}. The genitive (at least in the plural) can be analysed as suffixed to the oblique rather than to the nominative stem. The noun in the example is \textit{ṣinɡ} `horn':


\begin{table}[H]
\begin{tabular}{ l l l }
&
\textbf{Singular} &
\textbf{Plural} \\
Nominative &
\textit{ṣinɡ} &
\textit{ṣínɡ-a} \\
Oblique &
\textit{ṣínɡ-a} &
\textit{ṣínɡ-am} \\
Genitive &
\textit{ṣínɡ-ii} &
\textit{ṣínɡ-am"=ii} \\
\end{tabular}
\end{table}



There are three main functions of the oblique case of nouns: a) as the transitive subject in the Perfective; b) as the form to which postpostions are added; and c) as a~locative. A number of other case"=like functions (such as recipients) and more peripheral arguments appear as postpositional phrases.


Palula displays core"=case asymmetry \citep[206--209]{iggesen2005}, within the category of nouns as well as for NPs at large (more on pronouns below). While most nouns make a~nominative"=oblique distinction, one declensional class in particular does not make this distinction at all, whereas some of the pronouns make an~even more fine"=tuned nominative"=accusative"=ergative distinction:

\begin{table}[H]
\begin{tabular}{ l l | l | l }
&
\multicolumn{1}{l}{\textbf{`man'}} &
\multicolumn{1}{l}{\textbf{`woman'}} &
\textbf{\textbf{\textsc{3sg}}}\\
Nominative &
\textit{míiš} &
\textit{kúṛi} &
\textit{so} \\\cline{2-2}\cline{4-4}
Oblique/Accusative &
\textit{míiša} &
\textit{kúṛi} &
\textit{tas} \\\cline{4-4}
Ergative &
\textit{míiša} &
\textit{kúṛi} &
\textit{tíi} \\
\end{tabular}
\end{table}


Palula has a~fairly typical Indo"=European two"=gender system, which is primarily sex"=based \citep[130--133]{corbett2005a}. A noun is either masculine or feminine, a~property established through morphological agreement. The basis for gender assignment is semantic as well as formal \citep[134--137]{corbett2005b}.


Adjectives are inflected for agreement in gender (masculine, feminine), number (singular, plural) and case (nominative, non"=nominative). The adjective in the example is \textit{paṇáaru} `white':


\begin{table}[H]
\begin{tabular}{ l l l l }
&
\textbf{Masculine} &
&
\textbf{Feminine} \\
&
Singular &
Plural &
\\
Nominative &
\multicolumn{1}{l|}{\textit{paṇáaru}} &
\multicolumn{1}{l|}{\textit{paṇáara}} &
\textit{paṇéeri} \\\cline{2-2}
Non"=nominative &
\textit{paṇáara} &
\multicolumn{1}{ l|}{\textit{paṇáara}} &
\textit{paṇéeri} \\
\end{tabular}
\end{table}


Finite verbs are inflected for tense"=aspect, mood (in a~limited sense) and agreement in a)
gender/number, \textit{or} b) person (the type of agreement expressed depending on tense, see \sectref{sec:2-3}
below). There are also some nonfinite forms. For the sake of a~more economical presentation that
takes verbs of different inflectional classes into account, all verbs are analysed as having
a~perfective and a~non"=perfective stem. The verb in the example is \textit{til-} `walk':


\begin{table}[H]
\begin{tabular}{ l l l l }
&
&
\textbf{Singular} &
\textbf{Plural} \\
\textbf{Non"=perfective stem} &
&
&
\\
Present &
\textsc{m} &
\textit{til-áan-u} &
\textit{til-áan-a} \\
&
\textsc{f} &
\textit{til-éen-i} &
\textit{til-éen"=im} \\
Future &
1 &
\textit{tíl-um} &
\textit{til-íia} \\
&
2 &
\textit{tíl-aṛ} &
\textit{tíl-at} \\
&
3 &
\textit{tíl-a} &
\textit{tíl-an} \\
Imperative &
&
\textit{tíl} &
\textit{tíl-ooi} \\
Infinitive &
&
\textit{til-áai} &
\\
Converb &
&
\textit{til-í} &
\\
Obligative &
&
\textit{til"=eeṇḍeéu} &
\\
Copredicative Participle &
&
\textit{til-íim} &
\\
Verbal Noun &
&
\textit{til"=ainií} &
\\
Agentive Verbal Noun &
\textsc{m} &
\textit{til-áaṭ-u} &
\textit{til-áaṭ-a} \\
&
\textsc{f} &
\textit{til-éeṭ-i} &
\textit{til-éeṭ-im} \\
\textbf{Perfective stem} &
&
&
\\
Perfective/Perfective Participle &
\textsc{m} &
\textit{tilíl-u} &
\textit{tilíl-a} \\
&
\textsc{f} &
\textit{tilíl-i} &
\textit{tilíl-im} \\
\end{tabular}
\end{table}

The pronoun system proper (i.e. 1st and 2nd person) is interesting in that it makes more distinctions in the plural than in the singular:

\begin{table}[H]
\begin{tabular}{ l l l | l l }
&
\textbf{Nominative} &
\multicolumn{1}{l}{\textbf{Accusative}} &
\textbf{Genitive} &
\textbf{Ergative} \\
\textsc{1sg} &
\textit{ma} &
\textit{ma} &
\textit{míi} &
\textit{míi} \\\cline{2-5}
\textsc{2sg} &
\textit{tu} &
\textit{tu} &
\textit{thíi} &
\textit{thíi} \\\cline{2-5}
\textsc{1pl} &
\multicolumn{1}{ l|}{\textit{be}} &
\textit{asaám} &
\multicolumn{1}{ l|}{\textit{asíi}} &
\textit{asím} \\\cline{2-5}
\textsc{2pl} &
\multicolumn{1}{ l|}{\textit{tus}} &
\textit{tusaám} &
\multicolumn{1}{ l|}{\textit{tusíi}} &
\textit{tusím} \\
\end{tabular}
\end{table}


The forms \textit{ma} and \textit{tu} are glossed throughout this work as nominative and \textit{míi} and \textit{thíi} as genitive, regardless of their functions in the clause.


The demonstratives, which are used as third"=person pronouns, essentially make the same case distinctions as the plural personal pronouns. Additionally they display gender distinctions (in the nominative singular) as well as a~three"=way deictic contrast:


\begin{table}[H]
\begin{tabular}{ l l l l | l | l | l }
&
&
&
\textbf{Nominative} &
\textbf{Accusative} &
\textbf{Genitive} &
\textbf{Oblique} \\
Proximate &
\textsc{sg} &
\textsc{m} &
\textit{nu} &
\textit{nis} &
\textit{nisíi} &
\textit{níi} \\\cline{4-4}
&
&
\textsc{f} &
\textit{ni} &
\textit{nis} &
\textit{nisíi} &
\textit{níi} \\\cline{5-7}
&
\textsc{pl} &
&
\textit{ni} &
\textit{ninaám} &
\textit{niníi} &
\textit{niním} \\\cline{4-7}
Distal &
\textsc{sg} &
\textsc{m} &
\textit{lo} &
\textit{las} &
\textit{lasíi} &
\textit{líi} \\\cline{4-4}
&
&
\textsc{f} &
\textit{le} &
\textit{las} &
\textit{lasíi} &
\textit{líi} \\\cline{5-7}
&
\textsc{pl} &
&
\textit{le} &
\textit{lanaám} &
\textit{laníi} &
\textit{laním} \\\cline{4-7}
Remote &
\textsc{sg} &
\textsc{m} &
\textit{so} &
\textit{tas} &
\textit{tasíi} &
\textit{tíi} \\\cline{4-4}
&
&
\textsc{f} &
\textit{se} &
\textit{tas} &
\textit{tasíi} &
\textit{tíi} \\\cline{5-7}
&
\textsc{pl} &
&
\textit{se} &
\textit{tanaám} &
\textit{taníi} &
\textit{taním} \\
\end{tabular}
\end{table}


There is relatively little synchronically productive derivational morphology in the language, but a~productive process for deriving verbs from other categories from within the language as well as entirely from novel or non"=native elements is the use of verbalisers such as \textit{the-} `do' and \textit{bhe-} `become':


\begin{table}[H]
\begin{tabularx}{\textwidth}{ l l l l l l l}
\textit{madád} &
`help' &
+ &
\textit{the-} &
{\textgreater} &
\textit{madád thíilu} &
`helped'\\
\textit{tanɡ} &
`narrow' &
+ &
\textit{the-} &
{\textgreater} &
\textit{tanɡ thíilu} &
`troubled'\\
\textit{milaáu} &
`joined' &
+ &
\textit{bhe-} &
{\textgreater} &
\textit{milaáu bhílu} &
`met'\\
\textit{ašaq} &
`love' &
+ &
\textit{bhe-} &
{\textgreater} &
\textit{ašaq bhílu} &
`fell in love with'\\
\end{tabularx}
\end{table}



\section{Syntax}
\label{sec:2-3}
Word order in Palula is almost exclusively head"=final. This is seen in the word order in noun phrases (determiner"=noun, adjective"=noun, numeral"=noun, genitive"=noun), adjective phrases (adjunct"=adjective), and postpositional phrases (NP"=postpostion), as well as in clauses (SV and AOV), the latter of which will be obvious from most of the examples given below.


As far as alignment is concerned, Palula displays an~intricate split system. In the Perfective, the pattern is essentially ergative, seen in example (\ref{ex:2-1}) as a~non"=nominatively marked agent"=subject and verbal agreement with the feminine direct object, whereas in the non"=perfective categories, it is essentially accusative, which can be observed from the nominatively marked agent"=subject in (\ref{ex:2-2}), which is also the NP that the transitive verb agrees with in gender and number.

\begin{exe}
\ex
\label{ex:2-1}
\glll [íṇc̣-a] [čhéeli] khéel-i \\
bear[\textsc{msg}]-\textsc{ob} she.goat[\textsc{fsg}] eat.\textsc{pfv-}\textsc{f}\\
\textbf{A} \textbf{O} \\
\glt `The bear ate the goat.' (A:PAS056)
\end{exe}

\begin{exe}
\ex
\label{ex:2-2}
\glll [iṇc̣] áaṇc̣-a khiyáan-u\\
bear[\textsc{msg}] raspberry[\textsc{m}]-\textsc{pl} eat.\textsc{pres-m.sg} \\
\textbf{A} \textbf{O} \\
\glt `The bear is eating raspberries.' (A:KAT145)
\end{exe} 
Several NP splits further complicate the picture.


Agreement is part of all finite verb forms, but the particular agreement features realised are related to tense/aspect. In Future and Past Imperfective, the verb agrees with its target in person, as in (\ref{ex:2-3}), whereas in Present and the categories based on the Perfective, the verb agrees in gender and number, as can be seen in (\ref{ex:2-4}).

\begin{exe}
\ex
\label{ex:2-3}
\gll míi se preṣ-íi se bhraawú ma [ɡhaš-íin] de \\
\textsc{1sg.gn} \textsc{def} mother.in.law-\textsc{gn} \textsc{def} brother.\textsc{pl} \textsc{1sg.nom} catch-\textsc{3pl} \textsc{pst} \\
\glt `The brothers of that mother"=in"=law of mine captured me.' (A:HUA122)
\end{exe}

\begin{exe}
\ex
\label{ex:2-4}
\gll táapaṛ-a túuri íṇc̣-a čhéeli [ɡhašíl-i] hín-i \\ 
hill below bear-\textsc{ob} she.goat[\textsc{f}] catch.\textsc{pfv.fsg} be.\textsc{prs-f} \\
\glt `Below the hill, the bear had captured the goat.' (PAS054)
\end{exe}

Sentences lacking an~overt copula are allowed, and for predicate nominals in the Present tense, as the one shown in (\ref{ex:2-5}), they are normative.

\begin{exe}
\ex
\label{ex:2-5}
\gll míi baábu áak zamindaár míiš \\ 
\textsc{1sg.gn} father \textsc{idef} farmer man \\
\glt `My father is a~farmer.' (OUR002)
\end{exe}
Although it is possible to conjoin clauses with a~conjunctive suffix (also used for conjoining noun phrases), other strategies are preferred, such as juxtaposition for symmetrical clauses or the overwhelmingly favoured Converb construction, exemplified in (\ref{ex:2-6}), which is used for a~great variety of same"=subject clause combinations.

\begin{exe}
\ex
\label{ex:2-6}
\gll ti ba [bhun whay-í ba] [so mhaás muṭ-í bhun wheel-í ba] [teeṇíi ɡhooṣṭ-á the ɡhin-í] ɡáu \\
\textsc{3sg.ob} \textsc{prt} down come.down-\textsc{cv} \textsc{prt} \textsc{def.msg.nom} meat tree-\textsc{gn} down take.down-\textsc{cv} \textsc{prt} \textsc{refl} house-\textsc{ob} to take-\textsc{cv} go.\textsc{pfv.msg} \\
\glt `He came down [having come down], took down the meat from the tree [having taken down the meat from the tree], and brought it to his house.' (B:SHB762)
\end{exe}

In complex constructions, the unmarked order is a~complement clause followed by (or embedded in) the main clause (\ref{ex:2-7}), and similarly an~adverbial clause followed by (or, again, embedded in) the main clause (\ref{ex:2-8}). 

\begin{exe}
\ex
\label{ex:2-7}
\gll [neečíir theníi-a] díiš-a xalk-íim xwaaíš thíil-i \\
	hunt do.\textsc{vn"=pl} village\textsc{-ob} people\textsc{-pl.ob} desire do.\textsc{pfv-f} \\
\glt `People in the village wanted to go hunting.' (B:AVA200)
\end{exe}

\begin{exe}
\ex
\label{ex:2-8}
\gll[raaǰaá múṛ-u ta] putr-óom tasíi hukumát bulooṣṭéel-i \\
	king die.\textsc{pfv"=msg} \textsc{prt} son\textsc{-pl.ob} \textsc{3sg.gn} government snatch.\textsc{pfv-f} \\
\glt `When the king died, the sons seized the power.' (A:MAB003)
\end{exe}

However, a~post"=posed construction with the complementiser \textit{ki} is also commonly used (\ref{ex:2-9}), especially for utterance complements.

\begin{exe}
\ex
\label{ex:2-9}
\gll ɡhueeṇíi-am maníit-u ki [ni bíiḍ-a zinaawúr xálaka hín-a] \\
	Pashtun-\textsc{pl.ob} say.\textsc{pfv-m.sg} \textsc{Comp} \textsc{3pl.prox.nom} much-\textsc{mpl} wild people be.pres-\textsc{mpl} \\
\glt `The Pashtuns said: ``These are very wild people.''' (CHA008)
\end{exe}

Polar interrogatives are formed with a~clitical question particle \textit{-ee} (B. \textit{-aa}), as in (\ref{ex:2-10}), whereas an~indefinite"=interrogative pronoun (or other proform), such as \textit{kasée} (B.) `whose' in (\ref{ex:2-11}), is used in content interrogatives. 

\begin{exe}
\ex
\label{ex:2-10}
\gll ux-á díi hooǰóol-u ki tu insaán-ee \\
	camel-\textsc{ob} from ask.\textsc{pfv"=msg} \textsc{prt} \textsc{2sg.nom} human.being-\textsc{q} \\
\glt `He asked the camel: ``Are you a~man?''' (A:KIN007)
\end{exe}


\begin{exe}
\ex
\label{ex:2-11}
\gll aní [kasée] ziaarat-í thaní \\
	\textsc{prox.nom.pl} whose shrine-\textsc{pl} \textsc{qt} \\
\glt `Whose shrines are these?' (B:FOR026)
\end{exe}

Negation is formed with a~separate and invarable negative particle \textit{na}, preceding the predicate (\ref{ex:2-12}).

\begin{exe}
\ex
\label{ex:2-12}
\gll muṣṭúk-a xálak-a dhii-á díi [na] khooǰ-óon de \\
	of.past-\textsc{mpl} people-\textsc{pl} daughter-\textsc{ob} from \textsc{neg} ask-\textsc{3pl} \textsc{pst} \\
\glt `People in the old days were not asking their daughter [who she wanted to marry].' (A:MAR018)
\end{exe}
